---
title: "Reproducible Science & Figures Assignment"
output:
  html_document: default
  pdf_document: default
date: "2024-12-09"
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

*The following is a template .rmd RMarkdown file for you to use for your
homework submission.*

*Please Knit your .rmd to a PDF format or HTML and submit that with no
identifiers like your name.*

*To create a PDF, first install tinytex and load the package. Then press
the Knit arrow and select "Knit to PDF".*

## QUESTION 01: Data Visualisation for Science Communication

**Loading libraries**

```{r warning=FALSE, message=FALSE}
library(tidyverse)
library(palmerpenguins)
library(janitor)
library(ragg)
library(here)
library(svglite)
library(broom)
```

**Loading and saving raw data set**

```{r}
penguins_raw 
write_csv(penguins_raw, here("data", "penguins_raw.csv"))
```

**Loading functions**

```{r}
source(here("functions", "Functions_For_Penguins.R"))
```

**Cleaning the raw data and saving as a new data file**

```{r}
#clean data
penguins_cleaned <- penguins_raw %>%
  clean_column_names() %>%
  remove_columns("comments") %>%
  remove_columns("delta") %>%
  remove_empty_columns_rows() %>%
  shorten_species() %>%
  shorten_sex ()

#save and load clean data
write_csv(penguins_cleaned, here("data", "penguins_clean.csv"))

penguins_cleaned <- read_csv(here("data", "penguins_clean.csv"), show_col_types = FALSE)

#visualise the cleaned data
colnames(penguins_cleaned)
penguins_cleaned
```

**Creating a figure using the Palmer Penguin data set that is correct
but badly communicates the data**

```{r bad figure code, echo=FALSE}
# Saving dataset of species, flipper length & body mass columns and removing NA values 
species_flipper_length_and_bodymass <- penguins_cleaned %>% select(species, flipper_length_mm, body_mass_g) %>%
  drop_na()

# Visualising modified data set 
species_flipper_length_and_bodymass

# Saving modified data set to use in scatterplot
write_csv(penguins_cleaned, here("data", "species_flipper_length_and_bodymass.csv"))
```

#### a) Figure:

```{r echo=FALSE}
# Creating scatterplot
ggplot(species_flipper_length_and_bodymass, aes(x = flipper_length_mm, y = body_mass_g, color = species)) +
  geom_point(size = 10, alpha = 0.1) +  # Overly large points with excessive transparency
  labs(title = "Flipper Length vs Body Mass", 
       x = "Flipper Length (mm)", 
       y = "Body Mass (g)") +
  theme_minimal() +
  theme(axis.text = element_text(size = 3, angle = 90, color = "black"),
        legend.position = "top") +  
  scale_color_manual(values = c("pink", "yellow", "green")) +  
  coord_cartesian(xlim = c(150, 250), ylim = c(1000, 7000))

```

#### b) Explanation of poor design choices

This graph plots the body mass against flipper length for three penguin
species. However, the design choices in this graph are misleading and
poorly communicate the data. One reason for this is that the patterns in
the data are are difficult to see due to the overly large points with
excessive transparency that blend with one another, and are hard to see
over clashing colours. For example pink and yellow points overlap so we
cannot fully view the data for these two species. Additionally, the view
of the data is restricted by the axis as these do not start at 0, so any
potential outliers are not seen making the graph misleading. Finally,
the graphical elements of this graph are unclear as the text on the axis
is too small to read easily without zooming in, and the labels on the x
axis are oriented at a 90 degree angle making it awkward to read
(Whitlock and Schluter, 2019).

------------------------------------------------------------------------

## QUESTION 2: Data Pipeline

### Introduction

Sexual size dimorphism is a trait exhibited by many vertebrate species
as a result of selective forces acting differently on males and females
due to their differing reproductive and ecological roles. This trend has
been found to be true in penguins, particularly in differing flipper
lengths with males having larger flippers than females (Jennings et al.,
2016). This study will investigate sexual size dimorphism in the Adelie
species of penguin by testing whether males have larger flipper lengths
than females using. a linear model analysis of the 'Palmer Penguins'
data set. If there is no sexual size dimorphism in this species, we
would expect that the null hypothesis that there is no significant
difference in flipper length between males and females to be true . This
would mean that the coefficient of the linear model analysis would not
be statistically different from zero. If the coefficient of the linear
model analysis *is* statistically different from zero, we can reject
this null hypothesis and this would support the alternate hypothesis
that sexual size dimorphism in flippers does occur in Adelie penguins.

### Statistical Method

**Loading libraries**

```{r}
library(tidyverse)
library(palmerpenguins)
library(janitor)
library(ragg)
library(here)
library(svglite)
```

**Loading and saving raw data set**

```{r}
penguins_raw 
write_csv(penguins_raw, here("data", "penguins_raw.csv"))
```

**Loading functions**

```{r}
source(here("functions", "Functions_For_Penguins.R"))
source(here("functions", "Exploratory_Plotting.R"))
source(here("functions", "Diagnostic_Plot_Function.R"))
source(here("functions", "SaveAs_svg_png_Functions.R"))
```

**Cleaning the raw data and saving as a new data files**

```{r}
#clean data
penguins_cleaned <- penguins_raw %>%
  clean_column_names() %>%
  remove_columns("comments") %>%
  remove_columns("delta") %>%
  remove_empty_columns_rows() %>%
  shorten_species() %>%
  shorten_sex ()

#save and load clean data
write_csv(penguins_cleaned, here("data", "penguins_clean.csv"))

penguins_cleaned <- read_csv(here("data", "penguins_clean.csv"), show_col_types = FALSE)

#visualise the cleaned data
colnames(penguins_cleaned)
penguins_cleaned

# Saving dataset of Adelie penguins, flipper length & sex
adelie_flipper_length_and_sex <- penguins_cleaned %>%
  select(species, flipper_length_mm, sex) %>%
  filter(species == "Adelie") %>%
  drop_na()

write_csv(adelie_flipper_length_and_sex, here("data", "adelie_flipper_length_and_sex.csv"))
```

**Exploratory plot of raw data**

```{r Data Exploration}
# using exploratory plotting function generate violin plot
plot_flipper_violin(adelie_flipper_length_and_sex)

# Saving the plot as a PNG file in figures folder
save_flipper_violin_png(
  penguins_flippers = adelie_flipper_length_and_sex, 
  filename = "Figures/adelie_flipper_violin_plot.png", 
  size = 15,
  res = 300,
  scaling = 1
)

# Saving the plot as an SVG file in figures folder
save_flipper_plot_svg(
  penguins_flippers = adelie_flipper_length_and_sex,
  filename = "Figures/adelie_flipper_violin_plot.svg", 
  size = 15,
  scaling = 1
)

```

The violin plot above visualises the flipper length data for males and
females of the Adelie penguin species. From the shape of these plots, we
can see that the flipper length of males being on average larger than
that of the females. The data for both males and females appears
approximately normally distributed, with a slight positive skew for the
male flipper length data as the lower end of the violin is wider than
the top end.

**Defining the linear model**

```{r}
linear_model <- lm(flipper_length_mm ~ sex, data = adelie_flipper_length_and_sex)
```

**Checking data meets assumptions of linear model analysis**

Before conducting the linear model analysis we first verify that the
data meets the assumptions of a linear model analysis. These are;
independence of data points, normality, homoscedasticity, and absence of
influential outliers. We can assume that the data collection and
sampling methods used ensured independence of data points. To test for
compliance with the remaining assumptions, diagnostic plots are used to
examine the data.

```{r Diagnostic Plots}
#Diagnostic plots 
#Displaying diagnostic plots in a grid
par(mfrow = c(2, 2), 
    mar = c(4, 4, 2, 2))  


# Running diagnostic plot function
Diagnostic_Plot_Function()


# Saving diagnostic plots as a .png
save_diagnostic_plot_as_png(
  here("figures", "Diagnostic_Plot.png"),  
  Diagnostic_Plot_Function,                    
  size = 20,                               
  res = 300,                              
  scaling = 2                              
)

# Saving diagnostic plots as a .svg
save_diagnostic_plot_as_svg(
  here("figures", "Diagnostic_Plot.svg"),  
  Diagnostic_Plot_Function,                    
  size = 20,                               
  scaling = 2                              
)

```

The residuals versus fitted values and scale-location plot both show an
approximately horizontal red line which suggests the data fits the
assumption of homoscedasticity. The Q-Q residuals plot shows the points
falling mostly along the diagonal line with some deviation at the
extreme ends which are not sufficient to violate the assumption of
normality, therefore we may assume that the residuals are normally
distributed. Finally, the residuals versus factor plot indicates no
significant outliers, validating the assumption of absence of
influential outliers.

### Results & Discussion

**Summary of the linear model**

```{r Linear Model Summary}
# Viewing a summary of the linear model analysis
summary(linear_model)
```

**Plotted linear model**

```{r Plot of Linear model}
# Creating dataframe with the model confidence intervals 
model_predictions <- adelie_flipper_length_and_sex %>%
  mutate(
    predicted = predict(linear_model, newdata = adelie_flipper_length_and_sex),
    conf_low = predict(linear_model, newdata = adelie_flipper_length_and_sex, interval = "confidence")[,2],
    conf_high = predict(linear_model, newdata = adelie_flipper_length_and_sex, interval = "confidence")[,3]
  )

# Plotting the model
plot <- ggplot(model_predictions, aes(x = sex, y = flipper_length_mm)) + 
  geom_point(aes(color = sex), size = 2, alpha = 0.6) + 
  geom_errorbar(aes(ymin = conf_low, ymax = conf_high), width = 0.2) + 
  stat_summary(aes(x = sex, y = flipper_length_mm), fun = mean, geom = "point", color = "black", size = 2) + 
  labs(title = "Linear Model: Flipper Lengths in Male and Female Adelie Penguins", x = "Sex", y = "Flipper Length (mm)") +
  theme_minimal() + 
  theme(legend.position = "none") 

plot
```

**Saving Plots as .svg & .png files**

```{r Saving Plots}
# Saving as PNG
save_plot_as_png(plot, "figures/Linear_Model_Plot.png", size = 20, res = 300)

# Saving as SVG
save_plot_as_svg(plot, "figures/Linear_Model_Plot.svg", size = 20)
```

**Results table**

```{r Linear Model Results Table, echo=TRUE}

# Creating a summary of the linear model 
linear_model_summary <- tidy(linear_model)

# Creating row for female (intercept) estimate & standard error
female_row <- linear_model_summary %>%
  filter(term == "(Intercept)") %>%
  mutate(
    term = "Female", 
    estimate = estimate, 
    std.error = std.error
  ) %>%
  select(term, estimate, std.error)

# Creating row for male estimate, standard error, t statsitic and p value 
male_row <- linear_model_summary %>%
  filter(term == "sexMale") %>%
  select(term, estimate, std.error, statistic, p.value) 

# Forming the full table by combining rows and formatting for better visualisation
summary_table <- bind_rows(female_row, male_row)

summary_table <- summary_table %>%
  mutate(
    estimate = round(estimate, 3),
    std.error = round(std.error, 3),
    statistic = ifelse(is.na(statistic), NA, round(statistic, 3)),
    p.value = ifelse(is.na(p.value), NA, format(p.value, scientific = TRUE, digits = 4)),
    term = recode(term, "sexMale" = "Male", "sexFemale" = "Female")
  )

View(summary_table)

```

| Term   | Estimate  | Standard Error | T Statistic | P Value  |
|--------|-----------|----------------|-------------|----------|
| Female | 187.795   | 0.716          | NA          | NA       |
| Male   | 4.616     | 1.013          | 4.559       | 1.09e-05 |

From these results, the positive estimate for male Adelie penguins of
4.616 indicates that on average, males have a greater flipper length
than females by 4.616mm. Additionally, the standard errors are small
which indicates that the estimates are precise and the model is robust.
The P value of 1.09x10⁻⁵ is less than the accepted significance level
0.05, suggesting this difference in flipper length is statistically
significant and unlikely to have arisen by chance. Therefore, we have
sufficient evidence to reject the null hypothesis which stated no
difference in flipper length between males and females.

### Conclusion

The linear model analysis was conducted to look at the difference in
flipper length between male and female Adelie penguins on the basis that
this data appeared to meet the assumptions of a linear model. The
results of this model indicated a statistically significant relationship
at the 0.05 significance level (p-value of 1.09 × 10⁻⁵) with male
penguins on average having a greater flipper length than females. This
provides strong evidence for rejecting the null hypothesis that there is
no difference in flipper length between sexes. This difference in
flipper length may be influential in understanding the ecology or
evolution of sexual dimorphism in this species of penguin. Future
studies may endeavour to investigate other characteristic traits of
sexual dimorphism in Adelie penguins or compare this difference in
flipper length in other penguin species to determine whether they also
follow this trend.

### References

-   Jennings, S., Varsani, A., Dugger, K.M., Ballard, G. and Ainley,
    D.G. (2016). Sex-Based Differences in Adélie Penguin (Pygoscelis
    adeliae) Chick Growth Rates and Diet. *PLOS ONE*, 11(3), p.e0149090.
    doi:<https://doi.org/10.1371/journal.pone.0149090.>

-   Whitlock, M. and Schluter, D. (2009). *The Analysis of Biological
    Data*. Roberts Publishers.

------------------------------------------------------------------------

## QUESTION 3: Open Science

*GitHub link:*
<https://github.com/awesomeorangutan/Reproducible_Science>

*Partner's GitHub
link:*<https://github.com/The-Nedstar/ReproducableFigures>

### Reflection on your experience running partner's code 

Most of the code ran smoothly when running my partner's code, although
there were a couple of issues that needed amending. The first issue I
encountered was in the initialisation section at the beginning of the
code when loading the packages from my library as the 'plot3D' package
was unavailable. This appeared to be an issue with Apple Mac as this was
available on their windows device, highlighting a potential issue
with reproducibility on different devices. However, this package was
only necessary for their code in question 1 of the assignment, so after
deleting 'library(plot3D)' from the code this was fixed and there were
no further issues relating to the code in question 2. Within the code
for question 2, the comments and subheadings on each section of code
made it easy to follow and understand and the code all ran without any
issues. One change I would make is in improving the clarity of some
variable names such as 'penguins_MVF'. Upon first inspection of this it
wasn't completely clear what this data set was, so a more detailed name
such as \'Penguin_Mass_vs_Flippers\' could be an improvement to make
this easier to follow. The use of the graph-making function found in
\'penguins.R\' within the functions file was very efficient as it
created both the exploratory scatterplot graph and linear model graph
within one function through use of \'if\' commands. This and created
very clear graphs and it was easy to make small adjustments such as
changing the titles and labels on the graphs. This did have limitations
though, as deeper and more complex changes to the graphs were very
difficult to accomplish. One change I would like to have made is
cropping the x axis to remove blank space (as had been done for the y
axis) allowing the data points to fill a greater area of the graph,
rather than being clustered on the right-hand side. This amendment
involved altering the graph-making function, which was very difficult to
accomplish. Therefore, it may be better to have made two separate and
more simple functions for making the scatterplot and linear model plot
to improve the ease of manipulating graph aspects without breaking the
code. 

### Reflection on my code based on experience with partner's code and their review of mine

My partner made the following suggestions for my code:  

1) Make more use of functions, for example, a function could have been
used to form the male and female rows in creating the results table.  

I agree with this suggestion and this is something I would change in the
future, not only for creating the results table, but also potentially
making a function for plotting the linear model as this would allow it
to be used for analysing other variables in the dataset.  

2) Fix the issue of reproducibility in my exploratory violin plot
function. My function in \'Exploratory_Plotting.R\' made a violin plot
specifically for making violin plot of the flipper data for male and
female Adelie penguins. In this function, the variable are already
defined so it couldn\'t be used for other variables in the Palmer
penguins data set as you would have to alter the function.  

I agree with this point, as it raises a potential issue with
reproducibility if one wanted to use the function for visualising raw
data of other variables. It would have been better to make a more
generalised function that could be used in other contexts. This is
something I would change in the future.  

3) Variable names are too long and complicated. My partner highlighted
that some of the names of my variables are quite long, which makes the
code look over crowded and slightly messy.  

I agree with this to an extent, for example
\'adelie_flipper_length_and_sex\' is quite long. However, I believe this
to be an issue of personal preference. While I find that this helps my
code to be more easily understood, to others it may be slightly
convoluted. In future, I might think more about making my variable names
both concise and clear. An improved version I could have used
is \'adelie_flipper_and_sex\' as the only data collected on flippers is
their length, so including \'length\' in the variable name may be
unnecessary detailing.

Overall, through running my partner's code and my partner running my
code, I learned that when writing code for other people to use it is
important to make sure that sections of the code are properly annotated
and variable names are clear so that it is easy to understand and make
alterations. I also realised that its not only important for the main
code to be reproducible and run smoothly, but functions used within the
code should be reproducible and able to be applied to other variables in
the data set.
